A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-10 ISMB/ResponseToReviewers-Incarnation-ISMB-2013.tex

\Comment{Is the method available?}
\Answer{\textbf{FIXME: Do you prefer a redirection to the git-hub repo or should we setup a basic web server?}}

\subsection{Reviewer \#2}

\Comment{1) The $x^{\gc(s)}$ in equation (1) seems to me a novel contributing term that is of high importance. Perhaps its introduction should be highlighted more in the text.}
\Answer{We tried to make this point clearer in our revised version. However, it should be noted that this contribution is not a goal in itself, but rather a mean to perform stochastic sampling with respect to the weighted distribution. This distribution has all sorts of desirable mathematical properties, among which the fact that induced distributions for additive features (here the \GCContent) are nicely concentrated, allowing for a rejection-based approach.
\textbf{FIXME: Should we really says it's not a goal. This seems to diminish our work... I'll try to find a way to re-formulate that. I agree with the content though.}
}

\Comment{2) Concerning Table 2 in the Supplementary data, the way it is written (IncaRNAtion + RNAinverse) it looks like "glocal" search heuristic whereas in the table it seems local. Perhaps there is a small typo ambiguity that can be resolved.}
\Answer{Reviewer \#1 also pointed out some inconsistency/unclarity in our previous formulation. The second column indeed showed the final \GCContent of the glocal approach, following the postprocessing of sequences with a local-search approach. While such a postprocessing tends to reestablish \GCContent{}s that are  closer to 50\% \GC, this shift remains reasonable (at most ~6\% \GC). Furthermore, this shift is very predictable and offers the user the possibility to \emph{overshoot} for a more extreme \GCContent with the global approach, knowing that the postprocessing step will tend to average it.
\textbf{FIXME: We could eventually emphasize that we bring \RNAinverse in a region which is likely to satisfy the  MFE criterion and thus will require only minimal adjustments.}}

### 2013-03-10 ISMB/methods_ISMB.tex
bioinfo template
fix figs
Typo in equations + Formatting

The recursion must consider different cases, depending on whether the first position in $\Struct$ is:
\begin{description}
\item[{\bf Case 1.}] Unpaired ($\Struct = \ub \Struct'$):
\begin{align}
	\Z{\BoolTrue,\ub \Struct'}{a,b} = \Z{\BoolFalse,\ub \Struct'}{a,b}  &:=
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{\BoolFalse,\Struct'}{a',b}; 
\label{eq:Z_unpaired}
\end{align}
\item[{\bf Case 2.}] Paired to last position ($\Struct = \op \Struct' \cp$), stacking onto a pre-existing closing pair ($\N=\BoolTrue$):
\begin{align}
	\Z{\BoolTrue,\op \Struct' \cp}{a,b} &:=
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{\BoolTrue,\Struct'}{a',b'};
\label{eq:Z_stack}
\end{align}
\item[{\bf Case 3.}] Paired to some position ($\Struct = \op \Struct' \cp \Struct''$), but not stacking ($\N=\BoolFalse$ or $\Struct''\neq \varepsilon$):
\begin{align}
	\Z{\N,\op \Struct' \cp \Struct''}{a,b} &:=
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{\BoolFalse,\Struct'}{a',b'}
      \cdot\Z{\BoolTrue,\Struct''}{b',b}.
\label{eq:Z_rec}
\end{align}
\end{description}

@@

\begin{figure}
\resizebox{\textwidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Stochastic backtrack procedure for a given substructure $\Struct$: Either the first position is left unpaired (top), a base-pair is formed between the two extremities, stacking onto an exterior base-pair (middle), or paired without creating a stacking, defining two regions on which subsequent recursive calls are needed (bottom). For the empty structure (omitted here), the empty sequence is returned. Positions indicated in red are assigned at the current stage of the backtrack.\label{fig:stochastic}}
\end{figure}

\begin{algorithm}[t]
\DontPrintSemicolon
	\SetAlgoLined
\SetKwFunction{Backtrack}{StocBack}
\SetKwFunction{Random}{Random}
	rand $\leftarrow$ \Random$\left(\Z{\N,\Struct}{a,b}\right)$\tcp*[r]{Draw random number in $[0,\Z{\N,\Struct}{a,b}[$}
 \Switch{}{
   \lCase(\tcp*[f]{Empty structure $\Rightarrow$ empty sequence}){$\Struct=\varepsilon$}{\Return{$\varepsilon$}}
   \Case(\tcp*[f]{First position is unpaired}){$\Struct=\ub\, \Struct'$}{		
   \For{$a'\in\B$}{
			rand $\leftarrow$ rand $- x^{\gc(a')}\cdot \Z{\BoolFalse,\Struct'}{a',b}$\;
			\lIf{$\text{\rm rand}<0$}\Return{$a'.\Backtrack(a',b,\BoolFalse,\Struct')$}\;
  		}
   }
   \Case(\tcp*[f]{Extremities involved in stacking base pair}){$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}$  {\rm \bf and} $\N=\BoolTrue$}
   {		
			\For{$(a',b')\in\B\times\B$}
    {
				rand $\leftarrow$ rand $-
			 	x^{\gc(a'.b')}
			 	\cdot e^{{-\ES^{\beta}_{ab \to a'b'}}/{RT}}
			 	\cdot \Z{\BoolTrue,\Struct'}{a',b'}	$\;
				\lIf{$\text{\rm rand}<0$}{
					\Return{$a'.\Backtrack(a',b',\BoolTrue,\Struct'). b'$}\;		
				}
			}
		}
   \Other(\tcp*[f]{First position is paired without a stacking pair})
   {
				\tcp{$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}\,\Struct''$}
				\For{$(a',b')\in\B\times\B$}{
					rand $\leftarrow$ rand $-	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{\BoolFalse,\Struct'}{a',b'}
	    	  \cdot\Z{\BoolTrue,\Struct''}{b',b}$\;	
 					\lIf{{\rm rand} $<0$}{
						\Return{$a'
						.\Backtrack\left(a',b',\BoolFalse,\Struct'\right)
						.b'
						.\Backtrack\left(b',b,\BoolTrue,\Struct''\right)
						$}\;	
					}
				}
   }
 }
\caption{\protect\Backtrack$\left(a,b,\N,\Struct\right)$\label{alg:back}}
\end{algorithm}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been computed and memorized, a stochastic backtrack starts from the target structure $\Target$ with any exterior bases $[a,b]$ and no nesting base-pair, corresponding to a call \Backtrack$\left(\varnothing,\varnothing,\BoolFalse,\Target\right)$ to Algorithm~\ref{alg:back}.  At each step, a suitable assignment for one or several positions is chosen, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the appropriate substructures are then performed. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(n)$ time and space.

>>>>

The recursion must consider different cases, depending on whether the first position in $\Struct$ is:
\begin{itemize}
\item {\bf Case 1.} Unpaired ($\Struct = \ub \Struct'$):
\begin{align}
	\Z{\BoolTrue,\ub \Struct'}{a,b} = \Z{\BoolFalse,\ub \Struct'}{a,b}  &:=
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{\BoolFalse,\Struct'}{a',b}; 
\label{eq:Z_unpaired}
\end{align}
\item {\bf Case 2.} Paired to last position ($\Struct = \op \Struct' \cp$), stacking onto a pre-existing closing pair ($\N=\BoolTrue$):
\begin{align}
	\Z{\BoolTrue,\op \Struct' \cp}{a,b} &:=
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{\BoolTrue,\Struct'}{a',b'};
\label{eq:Z_stack}
\end{align}
\item {\bf Case 3.} Paired to some position ($\Struct = \op \Struct' \cp \Struct''$), but not stacking ($\N=\BoolFalse$ or $\Struct''\neq \varepsilon$):
\begin{align}
	\Z{\N,\op \Struct' \cp \Struct''}{a,b} &:=
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{\BoolTrue,\Struct'}{a',b'}
      \cdot\Z{\BoolFalse,\Struct''}{b',b}.
\label{eq:Z_rec}
\end{align}
\end{itemize}

@@

\begin{figure*}
\resizebox{\linewidth}{!}{\input{FigStochasticBacktrack.pgf}}
\caption{Stochastic backtrack procedure for a given substructure $\Struct$: Either the first position is left unpaired (top), a base-pair is formed between the two extremities, stacking onto an exterior base-pair (middle), or paired without creating a stacking, defining two regions on which subsequent recursive calls are needed (bottom). For the empty structure (omitted here), the empty sequence is returned. Positions indicated in red are assigned at the current stage of the backtrack.\label{fig:stochastic}}
\end{figure*}

\begin{algorithm}[t]
\DontPrintSemicolon
	\SetAlgoLined
\SetKwFunction{Backtrack}{SB}
\SetKwFunction{Random}{Random}
	rand $\leftarrow$ \Random$\left(\Z{\N,\Struct}{a,b}\right)$\tcp*[r]{Draw random number in $[0,\Z{\N,\Struct}{a,b}[$}
 \Switch{}{
   \lCase(\tcp*[f]{Empty structure $\Rightarrow$ empty sequence}){$\Struct=\varepsilon$}{\Return{$\varepsilon$}}
   \Case(\tcp*[f]{First position is unpaired}){$\Struct=\ub\, \Struct'$}{		
   \For{$a'\in\B$}{
			rand $\leftarrow$ rand $- x^{\gc(a')}\cdot \Z{\BoolFalse,\Struct'}{a',b}$\;
			\lIf{$\text{\rm rand}<0$}\Return{$a'.\Backtrack(a',b,\BoolFalse,\Struct')$}\;
  		}
   }
   \Case(\tcp*[f]{Extremities involved in stacking base pair}){$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}$  {\rm \bf and} $\N=\BoolTrue$}
   {		
			\For{$(a',b')\in\B\times\B$}
    {
				rand $\leftarrow$ rand $-
			 	x^{\gc(a'.b')}
			 	\cdot e^{{-\ES^{\beta}_{ab \to a'b'}}/{RT}}
			 	\cdot \Z{\BoolTrue,\Struct'}{a',b'}	$\;
				\lIf{$\text{\rm rand}<0$}{
					\Return{$a'.\Backtrack(a',b',\BoolTrue,\Struct'). b'$}\;		
				}
			}
		}
   \Other(\tcp*[f]{First position is paired without a stacking pair})
   {
				\tcp{$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}\,\Struct''$}
				\For{$(a',b')\in\B\times\B$}{
					rand $\leftarrow$ rand $-	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{\BoolFalse,\Struct'}{a',b'}
	    	  \cdot\Z{\BoolTrue,\Struct''}{b',b}$\;	
 					\lIf{{\rm rand} $<0$}{
						\Return{$a'
						.\Backtrack\left(a',b',\BoolFalse,\Struct'\right)
						.b'
						.\Backtrack\left(b',b,\BoolTrue,\Struct''\right)
						$}\;	
					}
				}
   }
 }
\caption{\protect\Backtrack$\left(a,b,\N,\Struct\right)$\label{alg:back}}
\end{algorithm}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been computed and memorized, a stochastic backtrack starts from the target structure $\Target$ with any exterior bases $[a,b]$ and no nesting base-pair, corresponding to a call \Backtrack$\left(\varnothing,\varnothing,\BoolFalse,\Target\right)$ to Algorithm~\ref{alg:back}.  At each step, a suitable assignment for one or several positions is chosen, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the appropriate substructures are then performed. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(n)$ time and space.
